---
title: "Class 12: Genome Informatics"
author: "Hyejeong Choi (PID: A16837133)"
format: pdf
toc: TRUE
---

# Section 1: : Identify genetic variants of interest

Downloaded CSV file from Ensemble https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39894595-39895595;v=rs8067378;vdb=variation;vf=959672880#373531_tablePanel

Here we read this CSV file

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")

head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

Find the percentage/proportion for each genotype

```{r}
round(table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100, 2)
```

Now let's look at a different population. I picked the GBR.

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")

head(gbr)
```

Find the proportion of G|G

```{r}
round(table(gbr$Genotype..forward.strand.) / nrow(gbr) *100, 2)
```

This variant that is associated with childhood asthma is more frequent in the GBR population than the MXL population.

Let's now dig into this further.

# Section 4: Population Scale Analysis

One sample is obviously not enough to know what is happening in a population. You are interested in assessing genetic differences on a population scale. So, you processed about ~230 samples and did the normalization on a genome level. Now, you want to find whether there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3 expression.

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")

head(expr)
```

```{r}
nrow(expr)
```

There are 462 samples total.

The sample sizes of each genotype:

```{r}
table(expr$geno)
```

The median expression for each genotype:

```{r}
expr_boxplot <- boxplot(exp ~ geno, expr)
```

```{r}
round(expr_boxplot$stats, 2)
```

The median for the A|A genotype is 31.25. The median for the A|G genotype is 25.06. The median for the G|G genotype is 20.07.

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
library(ggplot2)

ggplot(expr) +
  aes(x=geno, y=exp, fill=geno) +
  labs(x="Genotype", y="Gene Expression") +
  geom_boxplot(notch=TRUE)
```

The gene expression is higher on average with the A|A genotype than the G|G genotype. Yes, the SNP effects the expression of the ORMDL3 gene. Having a A|G genotype slightly increases the expression and the A|A genotype significantly increases the expression of the ORMDL3 gene. This shows that a SNP with the "G" allele decreases the gene expression.




